---
title: "Project 2 Report"
output: html_document
---

## Group Members

Specify the names and EIDs of all group members (if not working in a group just put your own name and EID):
Stephanie Kuo sk59344
Tony Huang tkh699
Annika Shivam aes5956 


## Question

State question being addressed and describe the outcome variable to be predicted. Explain why it is useful to answer this question.

**Can we predict whether an animal arrives at the Austin Animal Center in healthy (Normal) condition based on animal type, intake type, and sex upon intake?" The outcome variable is Intake Condition, recoded as a binary variable where 1 = Normal (healthy) and 0 = Unhealthy/Other conditions (Injured, Sick, Neonatal, Feral, etc.). This question is useful because predicting an animal's condition upon arrival can help the shelter better allocate medical resources and staff This can ensure animals that are likely to arrive in poor condition receive prompt attention and care.**


## Data and Setup

Load the dataset into R below. In addition, load any packages that you will need using the `library()` function. For example, you will need to load the `tidyverse` and `tidymodels` packages.


```{r}
## Add your code for reading in the data here

# Load packages
library(tidyverse)
library(tidymodels)
library(kknn)
library(plotROC)

# Load dataset
Animal_Intakes <- read_csv("Austin_Animal_Center_Intakes_20250407.csv.gz")

# Subsample to speed up computation
Animal_Intakes <- Animal_Intakes |> slice_sample(n = 20000)

# Recode outcome variable as binary
Animal_Intakes <- Animal_Intakes |>
    mutate(Healthy = factor(ifelse(`Intake Condition` == "Normal", 1, 0)))

# Take a look at the data
head(Animal_Intakes)

## If the dataset is too large for your computer to handle, you can draw a
## smaller random sub-sample of the data using the `slice_sample()` function. For
## example 
##
## dat_sub <- dat |> 
##     slice_sample(n = 50000) 
## 
## draws a random sample of 50,000 rows from the `dat` object.



```


## Base Prediction Model

Use either linear regression (for continuous outcome) or logistic regression (binary outcome) to build a base prediction model. Use the tidymodels framework to 

1. Split the dataset into a training and testing dataset using 75% of observations for training and 25% for testing; leave the test dataset aside for now.

2. Select variables to include in your model recipe. You may need to experiment with different variable combinations to find the best predicting model.

Use 10-fold cross-validation on the training data to estimate either 

* the root mean squared error (RMSE) if your outcome is continuous; or 
* the accuracy if your outcome is binary.


```{r}
# Set seed for reproducibility
set.seed(322)

# Split data into 75% training and 25% testing
intakes_split <- initial_split(Animal_Intakes, prop = 0.75)
train <- training(intakes_split)
test <- testing(intakes_split)

# Create recipe without Age upon Intake
rec <- train |>
    recipe(Healthy ~ `Animal Type` + `Intake Type` + `Sex upon Intake`)

# Create logistic regression model
model <- logistic_reg() |>
    set_engine("glm") |>
    set_mode("classification")

# Create workflow
wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

# Run 10-fold cross validation
folds <- vfold_cv(train, v = 10)
res <- fit_resamples(wf, resamples = folds)

# Show performance metrics
res |> collect_metrics()
```

Which configuration of variables in your model seems to produce the best performing model on the training data?

**The model using Animal Type, Intake Type, and Sex upon Intake as predictors produced the best performing model on the training data, achieving an accuracy of 85.20%.**


## Alternate Prediction Model

For this part of the analysis, choose one machine learning approach to build an alternate prediction model. The ones we have seen in class are K-Nearest Neighbors (`nearest_neighbor()`), decision trees (`decision_tree()`), and random forests (`rand_forest()`). 

1. Identify the tuning parameters for your chosen model.

2. Explore which combination of variables and tuning parameters produces good model performance.

3. Create a workflow that assembles your recipe and your model specification.

4. Use 10-fold cross-validation to estimate either the RMSE (continous outcome) or the accuracy (binary outcome) for your model in the training data.

5. Tune your model using a range of possible tuning parameters to identify the best-predicting model

NOTE: Doing the cross-validation and model tuning here may take a *significant* amount of time depending on your choice of tuning parameters and the speed of your computer. You may need to be patient with the process here.


```{r}
# KNN model specification with tuning parameter (k = number of neighbors)
knn_model <- nearest_neighbor(neighbors = tune()) |>
    set_engine("kknn") |>
    set_mode("classification")

# Use same recipe as before
knn_wf <- workflow() |>
    add_recipe(rec) |>
    add_model(knn_model)

# Define grid of k values to try (5 to 50, 10 values)
k_grid <- grid_regular(neighbors(range = c(5, 50)), levels = 10)

# Tune using the same 10-fold cross-validation folds from before
knn_tune <- tune_grid(
    knn_wf,
    resamples = folds,
    grid = k_grid,
    metrics = metric_set(accuracy)
)

# View all results
knn_tune |> collect_metrics()

# Find the best k
best_k <- knn_tune |> select_best(metric = "accuracy")
best_k

# Finalize workflow with best k
final_knn_wf <- knn_wf |> finalize_workflow(best_k)

```

Which one of your models (i.e. which combination of tuning parameters and variables) produces the best prediction performance on the training dataset?

**The KNN model with k = 30 produced the best performance among the KNN models tested, achieving a cross-validated accuracy of 83.96%. However, this is slightly lower than the base logistic regression model's accuracy of 85.20%, so the logistic regression remains the better performing model.**


## Final Model Fit

Choose the model that performs the best on your training dataset (either the linear/logistic regression or the machine learning model). Make a final assessment of your model using the testing dataset.


```{r}
# Fit final KNN model on the full training data and evaluate on test data
final_knn_fit <- wf |>
    last_fit(intakes_split)

# View final performance metrics on test data
final_knn_fit |> collect_metrics()

```


What are the final performance metrics for your best model?

**The final logistic regression model achieved an accuracy of 84.02% on the held-out test dataset, with a ROC AUC of 0.678. This is consistent with the cross-validated training accuracy of 85.20%, suggesting the model generalizes well to new data without significant overfitting.**



## Model Improvement


Suppose you had the ability to improve the fit of the model by collecting data on a new variable. What data would you most like to collect to improve the prediction performance of the model? Explain why you think collecting this new variable would improve the prediction performance. You may optionally include some code/analysis to support your explanation.

```{r}
## Optional: Add any supporting code here

```

**Collecting data on the environment the animal was in before transportation to the shelter could improve the fit of the model. If the animal is from a house, did there appear to be food and care, or was it largely neglected? If the animal is from a street, does it look like it's been struggling to find food? The classification of the environment quality could be as simple as very poor/poor/fair/good/excellent. Based off this classification, if an animal is from a fair or poor environment, it has a higher likelihood of needing more care and attention. This will help the shelter understand the health of the animal better, and will likely improve our model prediction as well.**




## Discussion

Reflect on the process of conducting this project. What was challenging, what have you learned from the process itself? Was there anything that was unexpected that you found during this analysis? If so, what was your expectation and how did the experience deviate from your expectation?

**We enjoyed reviewing concepts from the class and putting our models together to answer a question that could make a real impact. One of the difficult parts of this project was working to fine-tune the parameters for the model, but we learned that taking the time to do so ensures a more optimal outcome. We did not expect logistic regression model to outperform the KNN model, but for our question, it seems that a binary outcome was better predicted with our variables through a non machine-learning approach.**





## Submission to Gradescope

**Make sure to add your group members to the Gradescope submission!**